From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study

### 2025-05-16 draft_v1/draft_v1.tex
draft_update

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{amssymb}
\usepackage{booktabs} % Add to your preamble for cleaner table lines
\usepackage{makecell} % Add to your preamble for multi-line cells
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes.geometric}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causal Discovery on Precarity and Depression},
  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Causal Discovery on Precarity and Depression}

@@

\date{2025-02-11}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precariousness and
depression is critical for developing effective interventions. This
study utilizes data from the HELIUS cohort study to explore these
relationships using advanced causal discovery methods. By applying
algorithms such as FCI, and CCI, and combining traditional Gaussian CI
tests with non-parametric approaches like RCoT, we investigate how
different indicators of precariousness---including factors related to
employment, social relations and relational stress, financial situation,
and housing---affect depressed mood, both as a sum score and at the
individual symptom level. Our findings reveal that relational stress
consistently emerges as a potential causal factor for depressed mood,
while symptoms such as sleep disturbances, guilt, and anhedonia are
particularly sensitive to external stressors, acting as potential early
warning signals or intervention points for prevention. Moreover, the
results highlight complexities in the data, including the influence of
latent confounders and the challenges of capturing cyclic relationships.
Despite some limitations, such as unresolved ambiguities in causal
directions and challenges with mixed data distributions, this study
demonstrates the utility of causal discovery tools in disentangling the
intricate interplay between social and mental health dynamics. By
mapping these causal structures into computational models, future
research can simulate intervention effects, providing actionable
insights to mitigate the impact of precariousness on mental health. This
study serves as a foundational effort, offering both methodological
advancements and practical implications for addressing depression at a
population level.
\end{abstract}

@@

Mental health problems in urban areas are on the rise, yet the
complexity of mental health systems poses significant challenges in
understanding the underlying mechanisms driving these issues --- let
alone designing effective interventions. Research has long sought to
identify underlying factors contributing to mental health problems.
Recent research looked into the association between mental health and
indicators of precariousness in various dimensions of life, indicating a
high level of uncertainty and instability in people's lives, such as
factors related to employment, social relations, finances, housing, and
culture (ref Leonie's paper). This comprehensive perspective on
precariousness helped to highlight how different aspects of life may be
interconnected. While this research has advanced our understanding of
which indicators of precariousness are related to mental health, a key
question remains unanswered: how do these factors influence mental
health? Specifically, the lack of directional insight --- knowing what
influences what --- limits our ability to identify and prioritize
effective intervention targets.

This study aims to investigate the causal relationships between
different dimensions of precariousness and mental health outcomes, to
with a specific focus on depression, the most prevalent mental health
issue. Using causal discovery methods, we explore how various aspects of
precariousness contribute to depression and whether depression, in turn,
exacerbates precariousness, potentially creating a feedback loop.
Furthermore, we zoom in at the symptom level, examining individual
depressive symptoms to identify those that may act as initiators ---
triggering the activation of the symptom network --- by displaying
heightened sensitivity to specific precariousness factors. We also
investigate symptoms that may function as bridging elements, linking the
dynamics of depression with precariousness by feeding back into
precarious conditions, potentially reinforcing the cycle of mental
health deterioration and precarious conditions. Through this analysis,
we aim to uncover the causal mechanisms driving mental health challenges
related to precariousness and establish a foundation for more targeted
and effective interventions by identifying the key driving forces within
this system.

@@

We use data from the HELIUS study, a multi-ethnic cohort that includes
participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin
(Snijder et al., 2017). To operationalize precariousness factors, we
draw on the framework outlined in previous research (i.e., Leonie's
paper) and select a set of relevant variables.

To ensure a robust representation of each precariousness factor, we
conducted various exploratory analyses to identify consistent and
meaningful factor structures. Based on these analyses, we identified
five precariousness factors, including two related to recent stressors,
each comprising multiple variables as outlined below. Detailed
information on the exploratory analyses can be found in the
\hyperref[sec-appendix]{Appendix}.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay. \emph{P.emp} = employment precariousness; \emph{P.hou} =
housing precariousness; \emph{P.soc} = social precariousness;
\emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
relational stressors; \emph{Precarity} = overall precarity;
\emph{PHQsum} = PHQ-9 sum score; \emph{anh} = anhedonia; \emph{app} =
appetite; \emph{con} = concentration; \emph{dep} = depressed mood;
\emph{ene} = energy; \emph{glt} = guilty; \emph{mot} = motor; \emph{sui}
= suicidal}

@@

\subsection{Causal Discovery}\label{causal-discovery}

There are numerous causal discovery algorithms available; however, in
this study, we focus on algorithms suited to the potential cyclic
relationships within our system. Specifically, we use \emph{FCI} (Fast
Causal Inference) and \emph{CCI} (Cyclic Causal Inference), both capable
of accounting for such cycles (Mooij \& Claassen, 2020; Strobl, 2019).
Additionally, we include the \emph{PC} algorithm as a reference, given
its simplicity and prominence as one of the most widely known causal
discovery methods (Spirtes et al., 2001).

>>>>

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{amssymb}
\usepackage{booktabs} % Add to your preamble for cleaner table lines
\usepackage{makecell} % Add to your preamble for multi-line cells
\usepackage{tikz}
\usepackage{amsmath}
\usetikzlibrary{arrows.meta, shapes.geometric}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study},
  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{From Causal Discovery to Intervention Simulation: Modeling
Precarity and Depression in the HELIUS Study}

@@

\date{2025-05-14}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between financial
stress, domain-specific precariousness indicators, and depressive
symptoms. Our multi-resolution analysis---spanning both aggregate and
symptom-level constructs---identified financial stress as a robust
driver of depression, and highlighted key symptoms, such as sleep
disturbance and depressed mood, as likely initiators or reinforcers of
downstream precariousness. To explore how these structural insights
shape intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

Mental health disorders represent a growing global health concern,
especially in urbanized regions (Gruebner et al., 2017; World Health
Organization, 2022). In 2019, one in eight people worldwide were living
with a mental health condition, with the highest disability-adjusted
life years (DALYs) due to mental and addictive disorders concentrated in
high-income countries such as those in Northern Europe, North America,
and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
Despite growing awareness, governments have struggled to design
effective responses. Mental health outcomes arise from complex,
multi-level dynamics---intertwined with the social, economic, and
spatial structures of urban environments---making both diagnosis and
intervention design deeply challenging (Van Der Wal et al., 2021).

Recent work has shifted focus from individual-level vulnerability to
broader social and structural determinants. While high-income nations
show elevated overall DALYs, within-country disparities reveal that
income inequality is a significant predictor of mental health burden
(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
instability, and neighborhood disadvantage have all been linked to
increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
literature highlights that poor mental health is not merely a personal
issue, but one rooted in sustained socioeconomic stress and structural
uncertainty.

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains---including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025). This broader view reveals how different forms
of insecurity can co-occur and compound, forming an ecosystem of risk.
However, despite growing evidence of association between precariousness
and mental health, a fundamental question remains unresolved: \emph{How
do these factors causally influence one another?} Most existing studies
rely on cross-sectional correlations, leaving open questions about
directionality and feedback. Without stronger causal insight, it remains
difficult to identify priority targets for intervention or to anticipate
unintended effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precarity--depression relationship
affect the outcomes of stress-reducing interventions?} To answer this,
we combine two approaches. First, we use cycle-capable causal discovery
algorithms to infer directional relationships among financial stress,
domain-specific precariousness factors, and depressive symptoms. By
analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity
index) and disaggregated variables (individual symptoms and life-domain
indicators), we identify both broad associations and fine-grained causal
pathways. Second, we translate these insights into a computational
dynamical model that simulates how depression and precarity co-evolve
under varying levels of external stress. This model enables us to probe
how system properties---such as feedback strength and stochastic
noise---govern responsiveness to intervention. In some configurations,
the system exhibits bistability, remaining stuck in a high-risk state
unless external support exceeds a critical threshold. In others, it
shows unistable behavior, responding more gradually to incremental
changes. By integrating causal discovery with dynamic simulation, we aim
to demonstrate how the internal structure of the depression--precarity
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and which provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted various exploratory analyses to
identify consistent and meaningful data-driven factor structures. Based
on these analyses, we identified five precariousness factors, including
two related to recent stressors, each comprising multiple variables as
outlined below. DFurther details on the HELIUS study and factor
extraction are provided in the \hyperref[sec-appendix]{Appendix}.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay.\\
\emph{P.emp} = employment precariousness; \emph{P.hou} = housing
precariousness; \emph{P.soc} = social precariousness; \emph{S.fin} =
recent financial stressors; \emph{S.rel} = recent relational stressors;
\emph{Precarity} = overall precarity; \emph{PHQsum} = PHQ-9 sum score;
\emph{anh} = anhedonia; \emph{app} = appetite; \emph{con} =
concentration; \emph{dep} = depressed mood; \emph{ene} = energy;
\emph{glt} = guilty; \emph{mot} = motor; \emph{sui} = suicidal}

@@

\subsection{Causal Discovery}\label{sec-analysis}

To uncover directional relationships between precariousness and
depression, we used constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we
applied the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also included the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).

A key challenge in applying these algorithms to the HELIUS dataset is
that many variables are non-Gaussian and their relationships are likely
nonlinear. To address this, we supplemented the commonly used Gaussian
conditional independence (CI) test---based on partial
correlations---with a non-parametric alternative: the \emph{Randomized
Conditional Correlation Test} (RCoT). RCoT approximates kernel-based CI
testing using random Fourier features, significantly reducing
computational demands while maintaining sensitivity to nonlinear
dependencies (Strobl et al., 2019; Zhang et al., 2012).

We examined the causal structure using three complementary approaches:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Aggregate-level analysis: relationships between five domain-specific
  precariousness factors and the PHQ-9 sum score.
\item
  Fully disaggregated analysis: relationships between individual
  precariousness items and individual depressive symptoms.
\item
  Mixed analysis: relationships between individual symptoms and a
  composite index of overall precariousness.
\end{enumerate}

The aggregate analysis reduces dimensionality and yields interpretable
summaries of how domains of precarity relate to overall depression
severity. However, it may obscure finer-grained effects. The
disaggregated analysis allows for precise mapping of which symptoms are
influenced by (or influence) specific precariousness conditions, but it
introduces complexity due to the high dimensionality and distributional
properties of the data. The mixed approach, in turn, integrates these
perspectives by examining how individual symptoms respond to cumulative
precarity across domains.

To ensure robust estimation and guard against sensitivity to parameter
settings, we implemented a systematic grid of analysis conditions:

\begin{itemize}
\tightlist
\item
  Significance levels: \(\alpha\) = 0.01 and 0.05
\item
  Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
\item
  CI tests: Gaussian and RCoT
\item
  Algorithms: FCI, CCI, and PC
\end{itemize}

Each combination was bootstrapped 100 times for sum-score level analyses
and 30 times for symptom-level analyses, yielding a total of 1,600
graphs per algorithm for the full analysis. For symptom networks, we
fixed the skeleton structure using a consensus graph estimated across
PC, FCI, and CCI (\(\alpha\) = 0.01, RCoT). This constraint improves
computational efficiency without limiting the estimation of directional
edges between symptoms and precariousness variables.

To summarize the results, we adopted a two-level aggregation strategy.
First, within each analysis condition, we recorded the most frequently
observed edge-endpoint type (e.g., directed, undirected, uncertain) from
the bootstraps. Then, across all conditions, we tallied the most
frequent endpoint type per edge. In cases of ties, we assigned dashed
edges to reflect uncertainty. For transparency, we accompany each graph
with a table showing the relative frequencies of each edge symbol.

This procedure yields a final graph for each algorithm and analysis
type---offering both robustness and interpretability. See
Figure~\ref{fig-workflow} for an overview of the causal discovery
pipeline.

For detailed explanations of the algorithms, edge interpretations, and
graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions
and limitations, see the \hyperref[sec-appendix]{Appendix}.

\subsection{Modeling Dynamics between Precarity and
Depression}\label{modeling-dynamics-between-precarity-and-depression}

To explore how internal system structure shapes responses to external
support, we implement a high-level computational model that formalizes
key mechanisms suggested by the causal discovery analysis. Rather than
attempting to replicate the full estimated causal graph, this model
abstracts the essential structure into a compact, tractable system
focused on dynamic interactions between depression and social
precariousness, both influenced by financial stress.

As illustrated in Figure~\ref{fig-concept}, the model includes three
components: depression (\(D\)), represented by the PHQ-9 sum score,
social precariousness (\(P\)), and external financial stress (\(S\)).
Depression and precariousness are modeled as coupled state variables
that influence each other over time, with \(S\) acting as a shared
input. These dynamics are formalized as a pair of nonlinear stochastic
differential equations (SDEs):

\[
\begin{aligned}
dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
\end{aligned}
\]

Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
\(S\) is the external stressor associated with financial stress. The
parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
which variable adjusts toward its input-driven target state. The
coefficients \(\alpha_1, \alpha_2\) and \(\beta_1, \beta_2\) represent
the strengths of the causal influences between variables. Stochastic
fluctuations are introduced through the terms \(\sigma_1 dW_1\) and
\(\sigma_2 dW_2\), where \(dW_1\) and \(dW_2\) are independent Wiener
processes. The system uses the hyperbolic tangent function (\(tanh\)) to
bound the influence of inputs, ensuring smooth, saturating dynamics and
preventing runaway feedback loops. Importantly, the nonlinearity
introduced by the \(tanh\) function also allows the system to exhibit
bistability, supporting the existence of two stable states: one
corresponding to low depression and precariousness, and the other to
high depression and precariousness.

\includegraphics[width=0.35\textwidth,height=\textheight]{img/conceptmodel.png}

\caption{\label{fig-concept}Conceptual high-level causal structure of
precariousness and depression system.}

\subsubsection{Model Calibration}\label{model-calibration}

To identify realistic parameter settings for the simulation model, we
calibrated six free parameters---\(\alpha_1\), \(\alpha_2\),
\(\beta_1\), \(\beta_2\), and noise amplitudes \(\sigma_1\),
\(\sigma_2\)---using the Non-dominated Sorting Genetic Algorithm II
(NSGA-II) (Deb et al., 2002), a multi-objective evolutionary
optimization method well-suited for noisy and high-dimensional search
spaces. The goal was to match simulated system behavior to empirical
summary statistics derived from the HELIUS dataset, including the means
and variances of depression (\(D\)) and social precariousness (\(P\)),
as well as their partial correlations after adjusting for financial
stress (\(S\)), specifically \(\rho_{DP.S}\), \(\rho_{DS.P}\), and
\(\rho_{PS.D}\). Given the stochastic nature of the model, each
candidate parameter set was evaluated by simulating 1,000 independent
trajectories, each representing an individual with a fixed \(S\) value
sampled from the observed distribution. Final-state values were
extracted and aggregated to compute population-level summary statistics.
Fitness was assessed as the absolute deviation from the empirical
targets, aggregated across all objectives. NSGA-II was implemented via
the nsga2R package (Franz \& Nakamura, 2015), with a population of 300
evolved over 300 generations using standard hyperparameters (crossover
probability = 0.9, mutation probability = 0.25, and tournament size =
2). This process yielded a diverse ensemble of Pareto-efficient
solutions balancing trade-offs between objectives. From the resulting
Pareto front (Figure~\ref{fig-pareto}), we selected the top 30 parameter
sets with the lowest combined error to use in the subsequent simulations
assessing intervention effects.

\includegraphics[width=0.8\textwidth,height=\textheight]{img/paretoplot.png}

\caption{\label{fig-pareto}Pareto front of simulated parameter sets.
\emph{Each point represents a solution evaluated by its total variance
error (x-axis) and total partial correlation error (y-axis), with color
indicating total mean error. Red crosses highlight the top 30 solutions
used for simulation.}}

\subsubsection{Intervention Simulation}\label{intervention-simulation}

To evaluate the effects of hypothetical support, we simulate scenarios
in which each individual's financial stress level (\(S\)) is
proportionally reduced toward the minimum observed value. This models a
population-wide intervention applied at varying strengths --- from no
reduction in stress (0\%) to complete alignment with the minimum
observed stress level (100\%).

For each of the top 30 calibrated parameter sets, we simulate the
system's evolution across 300 individuals under a continuum of
intervention levels. Each simulation run generates final values for
\(D\) and \(P\), which are then used to assess how internal model
parameters influence the system's behavior under changing external
conditions.

\subsection{Causal Structure Linking Precarity and
Depression}\label{causal-structure-linking-precarity-and-depression}

\subsubsection{Depression as sum score}\label{depression-as-sum-score}

\subsubsection{Individual depression
symptom}\label{individual-depression-symptom}

\subsubsection{Precarity as sum score}\label{precarity-as-sum-score}

\subsection{Simulating the Dynamics Between Precarity and Depression
Under
Intervention}\label{simulating-the-dynamics-between-precarity-and-depression-under-intervention}

To evaluate how internal system structure influences the effect of
external support, we simulated outcomes under a range of intervention
strengths across the top 30 parameter sets from the calibrated model
ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
levels of depression (\(D\)) and precarity (\(P\)) after simulation,
with each line representing one parameter set. Lines are color-coded by
combined feedback and input strength, calculated as the product
\(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).

Parameter sets with stronger interactions exhibit more pronounced
declines in both depression and precarity in response to intervention.
At full intervention---modeled as a complete shift of external stress
(\(S\)) toward its observed minimum---the mean depression level
decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
These changes correspond to reductions of approximately 90.8\% relative
to their initial magnitudes, indicating that targeted intervention can
shift the system toward a substantially lower-risk equilibrium.

To better understand these dynamics, we focused on two illustrative
parameter sets: one with strong feedback/input coupling and low noise (a
``bistable'' candidate), and one with weaker coupling and higher noise
(a ``unistable'' candidate). In nonlinear systems, bistability refers to
the existence of two distinct stable equilibria---meaning the system's
long-term outcome depends on its starting point or perturbations.

In our model, bistability arises from the interaction between strong
internal coupling---particularly the product
\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
\(\tanh(\cdot)\) activation function saturates at high input values,
meaning sufficient gain is required to push the system into nonlinear
regimes where multiple stable points emerge. If the feedback gain is too
low, or the system is overwhelmed by noise, it behaves more linearly and
settles into a single equilibrium---yielding a unistable
configuration.\footnote{This resembles a classic \emph{pitchfork
  bifurcation} in discrete dynamical systems: when the gain \(k\) in a
  scalar map like \(x_{t+1} = \tanh(k x_t)\) exceeds a critical value,
  the system shifts from a single fixed point to two stable attractors
  and one unstable one. In our model, high values of
  \(\alpha_2 \times \beta_2\) similarly enable the emergence of multiple
  attractor states.}

The bottom panel of Figure~\ref{fig-intervention} illustrates these
contrasting behaviors. In the bistable model, a clear bifurcation
emerges: small interventions have little effect, but once a threshold is
crossed, the system abruptly transitions to a low-risk state. In the
unistable model, the transition is gradual and continuous, with outcomes
improving incrementally as stress is reduced.

These findings underscore that the impact of intervention depends not
only on its intensity, but on the system's internal configuration. In
bistable regimes, small or moderate interventions may be insufficient
unless they push the system past a tipping point. Unistable systems, by
contrast, respond more linearly, offering greater predictability and
flexibility in designing incremental interventions.

This distinction is especially relevant for mental health interventions
in populations exposed to persistent stressors like housing instability,
financial strain, or job insecurity. In such contexts, external
adversity may be reinforced through internal feedback loops between
psychological distress and precarious conditions. Recognizing when a
system is bistable may help identify when strong, early, or multifaceted
interventions are needed to produce lasting change---while unistable
dynamics may justify more gradual or targeted efforts.

Tailoring intervention strategies to the system's underlying structure
may enhance their effectiveness and sustainability, particularly in
addressing the compounding effects of socioeconomic precarity on mental
health.

\includegraphics[width=1\textwidth,height=\textheight]{img/intervention_plot2.pdf}

\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
precarity (\(P\)) across varying intervention strengths, simulated over
the top 30 parameter sets. Each line represents one parameter set and is
colored by its combined feedback and input strength.} \textbf{Bottom}:
Smoothed Density Heatmap of Final \(D\) and \(P\) for Two Parameter
Sets. \emph{One exhibits unistability (top two panels) and the other
exhibits bistability (bottom two panels). Bistable dynamics reveal a
sharp transition between distinct outcome regimes, while unistable
dynamics show gradual, continuous change across the intervention
gradient.}}

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precarity
interact---and how their internal structure modulates responsiveness to
intervention. Using cross-sectional data from the HELIUS study, we
applied cycle-capable discovery algorithms to estimate plausible
directional relationships among financial stress, multiple domains of
precariousness, and depressive symptoms. By analyzing both aggregate
constructs and individual variables, we aimed to capture patterns that
may be masked in either extreme of abstraction (i.e., purely aggregate
or disaggregated analyses).

The causal analysis consistently identified financial stress
(\emph{S.fin}) as a key plausible driver of depression, appearing across
both aggregated and disaggregated models. At the symptom level, we found
that sleep disturbance (\emph{slp}), depressed mood (\emph{dep}), guilt
(\emph{glt}), and anhedonia (\emph{anh}) were especially sensitive to
external stressors. Among these, \emph{dep} emerged as a central hub:
not only was it heavily influenced by other symptoms, it also exerted
directional influence on social precarity (\emph{P.soc}), suggesting a
feedback loop in which worsening mood contributes to social
disconnection, which in turn may exacerbate depressive symptoms. The CCI
algorithm even suggested a possible cyclic relationship between dep and
overall precarity, hinting at a structurally self-reinforcing mechanism.
\emph{Slp} also played a key initiating role. It showed directed
connections to both social and employment precarity and exhibited many
outgoing edges within the symptom network. This suggests \emph{slp} may
function as an early warning signal or entry point into the
depression--precarity cycle, making it a promising target for preventive
intervention. In contrast, \emph{P.soc} more often appeared as an effect
rather than a cause---receiving arrows from symptoms like \emph{dep} and
\emph{glt}, but rarely sending them. This asymmetry implies that social
precarity may often reflect downstream consequences of worsening mental
health rather than being its root cause. These patterns offer actionable
insights for intervention. Addressing symptoms like \emph{slp} early may
prevent escalation into broader depressive states or worsening
precarity. Intervening on \emph{dep}, as a central node with potential
feedback effects, could also disrupt the self-reinforcing dynamics
between mental health and social isolation.

To examine how such structures shape responsiveness to support, we
developed a simplified two-variable nonlinear dynamical model capturing
the feedback between depression and precarity, modulated by financial
stress. Calibrated against empirical summary statistics from the HELIUS
dataset, the model allowed us to simulate how these dynamics evolve
under varying levels of stress reduction. The simulations revealed that
intervention effectiveness depends not just on the magnitude of external
stressors, but on the internal configuration of the system itself. In
bistable regimes---characterized by strong feedback and low noise---the
system could remain stuck in a high-risk state unless an intervention
crosses a critical threshold, tipping it into a healthier equilibrium.
By contrast, unistable systems---driven by weaker coupling or greater
noise---responded more gradually and proportionally to increasing
support. These dynamics carry practical policy implications. In bistable
systems, even moderate interventions may have little effect unless they
are forceful enough to tip the system toward a lower-risk state. This
underscores the importance of early or intensive support when targeting
self-reinforcing cycles of depression and precarity. Unistable systems,
by contrast, are more responsive to incremental support, suggesting that
sustained, moderate interventions may be sufficient in such contexts.

That said, our study has several limitations. While the FCI and CCI
algorithms are designed to recover causal structure from cross-sectional
data---including feedback and latent confounding---many edge directions
in our results remain ambiguous, as indicated by the prevalence of
circle endpoints and bidirectional arrows, especially in weakly
connected regions of the graph. These uncertainties were most evident in
the links between individual symptoms and precariousness factors, where
edge estimates varied across algorithms and conditional independence
(CI) tests. This likely reflects limited statistical signal in
high-dimensional settings, where large conditioning sets reduce power
and small effect sizes are harder to detect. Although the CCI algorithm
is explicitly designed to detect cycles, we found little clear evidence
of feedback: most candidate loops appeared only as bidirectional edges,
which may reflect latent confounding or statistical imprecision rather
than true cyclic structure. This could point to either an absence of
strong feedback at the symptom level or limitations in the sensitivity
of full-graph discovery approaches applied to densely interconnected
variables. More focused strategies, such as local structure tests or
theory-guided subgraph modeling, may help resolve these ambiguities. In
addition, while our analyses relied solely on observational data, some
structural uncertainties---particularly those involving latent
confounding or bidirectional influence---may ultimately require
interventional or semi-interventional designs. Extensions such as LLC
(Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023), or
Bicycle (Rohbeck et al., 2024), which integrate observational and
interventional data, may prove valuable where such data are available.

Second, our findings showed sensitivity to the choice of conditional
independence (CI) test. The Gaussian CI test, which assumes linear
relationships and Gaussian noise, often produced denser graphs---likely
due to its tendency to overestimate weak or noisy linear associations.
In contrast, RCoT, a nonparametric alternative, avoids these assumptions
but relies on a Gaussian radial basis function (RBF) kernel to assess
dependence. While flexible in capturing nonlinear relationships, this
kernel is optimized for smooth, continuous variables and may
underperform with discrete or ordinal data. Its similarity estimates can
be unreliable when applied to variables with discontinuities,
potentially leading to missed dependencies---particularly relevant in
the HELIUS dataset, where some variables are ordinal or mixed-type
(Howlett, 2001). To address this, future work should consider adapting
kernel-based CI tests to better handle mixed data structures---for
example, by employing hybrid kernels that explicitly combine continuous
and discrete similarity measures. This could improve the reliability of
nonparametric causal discovery in settings with heterogeneous variable
types.

Finally, the structure of the dynamical model introduces its own
limitations. While the model was designed to reflect key features of the
discovered causal structure, it represents a highly simplified system:
only two variables---depression and social precarity---modulated by
external financial stress. This abstraction allowed for tractable
analysis and visual intuition, particularly around nonlinear behaviors
such as bistability, but omits other potentially influential constructs
highlighted in the causal graphs, including relational stress,
employment conditions and housing conditions. Moreover, the model's
nonlinearity---introduced via the hyperbolic tangent function---was
chosen for convenience rather than empirical justification. While this
function supports saturation and the emergence of tipping points, the
presence of bistability is a product of simulation, not empirical
verification. It remains possible that simpler, linear dynamics could
account for the observed data just as well. Future work should further
investigate whether bistability is a necessary feature of
depression--precarity dynamics, and explore alternative functional
forms---both linear and nonlinear---using richer data and stronger
empirical constraints.

Our results reveal that the effectiveness of interventions targeting
depression and precarity is shaped not only by the strength of external
stressors but also by the internal structure of the system. The
interplay between feedback strength and noise determines whether a
system responds gradually or only after crossing a tipping
point---highlighting the need for intervention strategies attuned to
these underlying dynamics.

These findings underscore the importance of considering internal
feedback---not just external adversity---when designing interventions.
Mental health systems are not always passively responsive to support;
their structure can constrain how much impact any given policy has. By
diagnosing these structural features---through tools like causal
discovery and computational modeling---researchers and policymakers can
better tailor interventions to the underlying mechanics of
vulnerability.

Beyond these specific findings, this study offers a generalizable
framework for studying complex mental health phenomena. Causal discovery
enables principled hypothesis generation from observational data, while
dynamical modeling helps test how those structures behave under
intervention. Together, these methods move us from descriptive patterns
to mechanistic insight, with practical relevance for real-world mental
health policy.

Future work could build richer dynamical models grounded in
higher-resolution empirical data---potentially incorporating time,
interventions, or more granular psychosocial constructs. Such extensions
would help validate whether mechanisms proposed here, such as
bistability, manifest in real-world trajectories. Ultimately, we hope
this study encourages further integration of causal structure learning
and system-level modeling to improve mental health outcomes under
conditions of socioeconomic precarity.

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-deb2002fast}
Deb, K., Pratap, A., Agarwal, S., \& Meyarivan, T. (2002). A fast and
elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
Transactions on Evolutionary Computation}, \emph{6}(2), 182--197.

\bibitem[\citeproctext]{ref-dojer2016learning}
Dojer, N. (2016). Learning bayesian networks from datasets joining
continuous and discrete variables. \emph{International Journal of
Approximate Reasoning}, \emph{78}, 116--124.

\bibitem[\citeproctext]{ref-elsenburg2025clustering}
Elsenburg, L. K., Nicolaou, M., Galenkamp, H., Lakerveld, J., \&
Stronks, K. (2025). The clustering of disadvantage in different life
dimensions across ethnic groups: A network analysis of indicators of
precariousness in the HELIUS study. \emph{Social Science \& Medicine},
117970.

\bibitem[\citeproctext]{ref-fone2014effect}
Fone, D., White, J., Farewell, D., Kelly, M., John, G., Lloyd, K.,
Williams, G., \& Dunstan, F. (2014). Effect of neighbourhood deprivation
and social cohesion on mental health inequality: A multilevel
population-based longitudinal study. \emph{Psychological Medicine},
\emph{44}(11), 2449--2460.

\bibitem[\citeproctext]{ref-nsga2R}
Franz, R., \& Nakamura, R. (2015). \emph{nsga2R: Elitist non-dominated
sorting genetic algorithm for multi-objective optimization}.
\url{https://CRAN.R-project.org/package=nsga2R}

\bibitem[\citeproctext]{ref-galenkamp2017measurement}
Galenkamp, H., Stronks, K., Snijder, M. B., \& Derks, E. M. (2017).
Measurement invariance testing of the PHQ-9 in a multi-ethnic population
in europe: The HELIUS study. \emph{BMC Psychiatry}, \emph{17}, 1--14.

\bibitem[\citeproctext]{ref-gruebner2017cities}
Gruebner, O., Rapp, M. A., Adli, M., Kluge, U., Galea, S., \& Heinz, A.
(2017). Cities and mental health. \emph{Deutsches {Ä}rzteblatt
International}, \emph{114}(8), 121.

\bibitem[\citeproctext]{ref-pevalin2017impact}
Pevalin, D. J., Reeves, A., Baker, E., \& Bentley, R. (2017). The impact
of persistent poor housing conditions on mental health: A longitudinal
population-based study. \emph{Preventive Medicine}, \emph{105},
304--310.
